Synthesis of Linear Functional Programs

### 2021-05-08 article/article.tex
first related work small

The simply-typed $\lambda$-calculus is a typed core functional
language~\cite{} that captures the essence of a type system in a simple and familiar environment. Its syntax consists of
functional abstraction, written $\lambda x{:}A.M$, denoting
an (anonymous) function that takes an argument of type $A$, bound to
$x$ in $M$; and application $M\,N$, with the standard meaning, and
variables $x$. For instance, $\lambda x{:}A.x$, the identity function, is a functional abstraction
taking an argument of type $A$ and returning it back.

\mypara{Propositions as Types}
%
It turns out that the inference rules of the simply-typed
$\lambda$-calculus are closely related to those of a system of natural
deduction for intuitionistic logic~\cite{prawitznd65}. This
relationship, known as the Curry-Howard correspondence~\cite{},
identifies that the propositions of intuitionistic logic can be read
as types for the simply-typed $\lambda$-calculus (``propositions as types'' observation), their proofs are
exactly the program with that type (``proofs as programs'' observation), and checking a proof is type checking a program (``proof checking as type checking'' observation)~\cite{}.

@@

We introduce a few typing rules to display how a type system can be
formalized. The following rule captures the nature of a
hypothetical judgment, allowing for reasoning from assumptions:
\[
    \infer*[right=(var)]
    {  }
    {\Gamma, u: A \vdash u : A}
\]
When seen as a typing rule for the $\lambda$-calculus, it
corresponds to the rule for typing variables -- variable $u$ has
type $A$ if the typing environment contains a variable $u$ of
type $A$.

@@

Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
resource-aware logic, where propositions are interpreted as resources
that are consumed during the inference process.  Where in standard
propositional logic we are able to use an assumption as many times as
we want, in linear logic every resource (i.e., every assumption) must
be used \emph{exactly once}, or \emph{linearly}. This usage
restriction gives rise to new logical connectives, based on the way
the ambient resources are used. For instance, conjunction, usually
written as $A\wedge B$, appears in two forms in linear logic:
multiplicative or simultaneous conjunction (written $A\tensor B$); and
additive or alternative conjunction (written $A\with
B$). Multiplicative conjunction denotes the simultaneous availability
of resources $A$ and $B$, requiring both of them to be
used. Alternative conjunction denotes the availability of $A$ and $B$,
but where only one of the two resources may be used. Similarly,
implication becomes into linear implication, written $A\lolli B$,
denoting a resource that will consume (exactly one) resource $A$ to
produce a resource $B$.
To present the formalization of this logic, besides the new connectives, we 
need to introduce the \emph{resource-aware context} $\Delta$.
 In contrast to the previously seen $\Gamma$, $\Delta$ is also a list of variables and their types,
 but in which each and every variable must be used exactly once.
To introduce the connective $\tensor$ which defines a multiplicative pair of propositions,
 we must use exactly all the resources ($\Delta_1, \Delta_2$) needed
 to make the ($\Delta_1$) first proposition, and ($\Delta_2$) second proposition:
\[
    \infer*[right=($\tensor I$)]
    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
\]
Out of the logical connectives, we need to mention one more, since it'll alter the judgment form, and it's the one 
that ensures logical strength -- i.e. we're able to translate usual logic into linear logic.
The $\bang$ is used (under certain conditions) to make a variable ``infinite'' i.e. to make
it useable more than once. To distinguish the ``infinite'' variables, a separate context is used --
$\Gamma$. $\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources that can only be used once.
The linear typing judgment for the introduction of the exponential $\bang$ takes the form:
\todo{A formula está um pouco estranha (faltam espaços)?}
\[
    \infer*[right=($\bang I$)]
    {\Gamma ; \emptyset \vdash M : A}
    {\Gamma ; \emptyset \vdash !M : !A}
\]
\todo{Describe exponent introduction and new gamma context}
\todo{Add h spaces in formula}
\[
    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta_1 \vdash M : !A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
    {\Gamma ; \Delta_1, \Delta_2 \vdash let !u = M \mathsf{in} N : C}
\]
\todo{Describe exponent elimination and new gamma context}

@@

\mypara{Type-and-Example-Directed Program Synthesis}~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
In this paper, type synthesis is explored

>>>>

The simply-typed $\lambda$-calculus is a typed core functional
language~\cite{} that captures the essence of a type system in a simple and familiar environment. Its syntax consists of
functional abstraction, written $\lambda x{:}A.M$, denoting
an (anonymous) function that takes an argument of type $A$, bound to
$x$ in $M$; and application $M\,N$, with the standard meaning, and
variables $x$. For instance, the term $\lambda x{:}A.x$, denoting the identity function, is a functional abstraction
taking an argument of type $A$ and returning it back.

\mypara{Propositions as Types}
%
It turns out that the inference rules of the simply-typed
$\lambda$-calculus are closely related to those of a system of natural
deduction for intuitionistic logic~\cite{prawitznd65}. This
relationship, known as the Curry-Howard correspondence~\cite{},
identifies that the propositions of intuitionistic logic can be read
as types for the simply-typed $\lambda$-calculus (``propositions as types'' observation),
their proofs are exactly the program with the given type (``proofs as programs'' observation),
and checking a proof is type checking a program (``proof checking as type checking'' observation)~\cite{}.

@@

We introduce a few typing rules to both show how a type system can be
formalized and to show the relationship with (intuitionistic)
propositional logic. The following rule captures the nature of a
hypothetical judgment, allowing for reasoning from assumptions:
\[
    \infer*[right=(var)]
    {  }
    {\Gamma, u: A \vdash u : A}
\]
When seen as a typing rule for the $\lambda$-calculus, it
corresponds to the rule for typing variables -- variable $u$ has
type $A$ if the typing environment contains a variable $u$ of
type $A$.

@@

Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
resource-aware logic, where propositions are interpreted as resources
that are consumed during the inference process.  Where in standard
propositional logic we are able to use an assumption as many times as
we want, in linear logic every resource (i.e., every assumption) must
be used \emph{exactly once}, or \emph{linearly}. This usage
restriction gives rise to new logical connectives, based on the way
the ambient resources are used. For instance, conjunction, usually
written as $A\wedge B$, appears in two forms in linear logic:
multiplicative or simultaneous conjunction (written $A\tensor B$); and
additive or alternative conjunction (written $A\with
B$). Multiplicative conjunction denotes the simultaneous availability
of resources $A$ and $B$, requiring both of them to be
used. Alternative conjunction denotes the availability of $A$ and $B$,
but where only one of the two resources may be used. Similarly,
implication becomes into linear implication, written $A\lolli B$,
denoting a resource that will consume (exactly one) resource $A$ to
produce a resource $B$.

To present the formalization of this logic, besides the new
connectives, we need to introduce the \emph{resource-aware context}
$\Delta$.  In contrast to the previously seen $\Gamma$, $\Delta$ is
also a list of variables and their types, but where each and every
variable must be used exactly once during inference.  So, to introduce
the connective $\tensor$ which defines a multiplicative pair of
propositions, we must use exactly all the resources
($\Delta_1, \Delta_2$) needed to realize the ($\Delta_1$) 
proposition $A$, and ($\Delta_2$) proposition $B$:
\[
    \infer*[right=($\tensor I$)]
    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
\]
Out of the logical connectives, we need to mention one more, since
it'll alter the judgment form, and it's the one that ensures logical
strength -- i.e. we're able to translate intuitionistic logic into
linear logic.
  The proposition $\bang A$ (read \emph{of course} $A$)
is used (under certain conditions) to make a resource ``infinite''
i.e. to make it useable an arbitrary number of times. To distinguish the
``infinite'' variables, a separate, unrestricted, context is used -- $\Gamma$. So
$\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources
that can only be used once.
The linear typing judgment for the introduction of the exponential $\bang$ takes the form:
\todo{A formula está um pouco estranha (faltam espaços)?}
\[
    \infer*[right=($\bang I$)]
    {\Gamma ; \emptyset \vdash M : A}
    {\Gamma ; \emptyset \vdash !M : !A}
\]
\todo{Describe exponent introduction and new gamma context}
\todo{Add h spaces in formula}
\[
    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta_1 \vdash M : !A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
    {\Gamma ; \Delta_1, \Delta_2 \vdash let !u = M \mathsf{in} N : C}
\]
\todo{Describe exponent elimination and new gamma context}

@@

\mypara{Type-and-Example-Directed Program Synthesis}~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
This paper explores a purely functional, recursive program synthesis technique based on types/proof search (as we intend to do),
with examples as an auxiliar technique to trim down the program synthesis vast search space.
\todo[inline]{more introduction to the paper}.
A good use is made of the general strategy of using a type system
to generate programs (by ``inverting'' it), rather than to type-check one.
Moreover, the paper uses a data structure called ``refinement tree'' in conjunction
with the extra information provided by the examples to allow for efficient synthesis
of non-trivial programs.
